Chemical tagging can work: \\ Identification of stellar phase-space structures \\ purely by chemical-abundance similarity

### 2016-01-18 papers/clusters.tex
typo
caption
kmeans text expansion badge unlocked

Part of the reason that chemical tagging still remains unrealized
is because the level of abundance specificity required
is very high.
If there are thousands of (relevant) molecular clouds forming stars in
the recent history of the Milky Way, clumps of stars can only stand
out in a high-dimensional abundance space: therefore accurate, or at
least precise, measurements of many different abundances are needed
for stellar sibling to have sufficiently unique fingerprints.
Stellar spectroscopic surveys now have the resolution, signal-to-noise,
wavelength coverage, and sample sizes to deliver many different chemical tags for each star.

% ARC to DWH:   "stand out in a high-dimensional abundance space" -->
%		 	    They'll only stand out in a high-dimensional abundance
% 				space, or a low-dimensional abundance space where those
% 				dimensions are all very orthogonal (e.g., different
% 				nucleosynthetic pathways). Should we add anything to that
%				effect?

@@

ARC asks DWH: Why aren't we showing The Cannon Teff/logg in these
				figures? Even if they aren't used for clustering..

@@

The k-means algorithm has disadvantages, of course.
One is that $K$ must be chosen by hand (or heuristically at best).
Here we are only demonstrating a concept; we don't need to have the
best possible clustering.
For this reason we simply choose $K=128$, $K=256$, and $K=512$ and look at all the
results.
The algorithm also only performs local optimization.
At each $K$ we perform 32 restarts with different initializations, and
preserve the best clustering (best according to the k-means score).

% ARC asks: how did you chose the initializations?

@@

DWH: EXPLICITLY GIVE THE ABUNDANCE-SPACE SCALES USED.

@@

When the k-means results are returned, the $D\times D$ empirical
variance tensor for the members of each cluster can be constructed,
and an effective density can be computed as the number of points in
the cluster divided by the square-root of the determinant of the
tensor.
This density was used to rank abundance-space overdensities (from highest
to lowest) for visual inspection.
In \figurename~\ref{fig:densities} we show the distribution of cluster membership
and density for the $K=256$ run of k-means.

We chose a few interesting cases from the high-density clusters in the $K=256$ run and show them in
\figurename~\ref{fig:M13}, \ref{fig:Pal5}, \ref{fig:Sgr},
\ref{fig:halo}, and \ref{fig:disk}.
The first three of these are dominated by stars in the known clusters M13, Pal5, and Sagittarius.
The fourth is a halo structure with high velocitay dispersion and possibly accreted.
The fifth is a thin-disk star-formation feature.
We will discuss these more below.

DWH: The clusters shown in \figurename~\ref{fig:M13}, \ref{fig:Pal5},
and \ref{fig:Sgr} might not look dense in abundance space, but they
are very dense in the 15-dimensional space.  Indeed,
\figurename~\ref{fig:M13} is the densest cluster found in the 15-d
space \emph{by far} (see \figurename~\ref{fig:densities}).

DWH: The figures show results only from the $K=256$ run, why?

DWH: Even if $K$ was chosen very carefully, it would still break up
real clusters into many clusters, why?  Abundance spreads!

DWH: The algorithm sees many different clusters; the figures show only
a few.  Do we see all clusters we \emph{should} see?  That question is
beyond the scope of this \documentname.

DWH: Each cluster figure and the disk figure shows stars that are part
of the structure and some that are not.  Every cluster will have
background contamination, why?

\begin{figure}[!p]
\includegraphics[width=2\figurewidth]{./figs/clusters_0256.pdf}
\caption{The distribution of membership and density for the 256
  abundance-space clusters returned by the k-means algorithm at
  $K=256$.  The densest cluster---which is shown in
  \figurename~\ref{fig:M13}---is far more dense than the second
  densest.\label{fig:densities}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0253}
\caption{Same as \figurename~\ref{fig:all} except that the entire
  sample has been made gray and the members of the most dense
  $K=256$ abundance-space clusters have been rendered as unique, prominent triangles.
  This cluster, which was identified only in abundance space (six
  projections of which are the top six panels in this \figurename),
  turns out to be dominated by the halo globular cluster
  M13.\label{fig:M13}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0034}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by the disrupting globular
  cluster Palomar 5.\label{fig:Pal5}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0177}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by the Sagittarius star
  cloud.\label{fig:Sgr}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0010}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a high-velocity-dispersion
  structure in the Galaxy halo.\label{fig:halo}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0141}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a very thin young stellar
  structure in the Galaxy disk.\label{fig:disk}}
\end{figure}

>>>>

Part of the reason that chemical tagging still remains unrealized
is because the level of abundance specificity required
is very high.
If there are thousands of (relevant) molecular clouds forming stars in
the recent history of the Milky Way, clumps of stars can only be
identified in abundance space if abundance space is high in dimensionality.
(In principle, it needs to be high in dimensionality both in terms
of the number of measured abundances \emph{and} in terms of the number of
neucleosynthetic pathways or the dimensionality of the true abundance
space.)
Therefore accurate---or at
least precise---measurements of many different abundances are needed
for stellar sibling to have sufficiently unique fingerprints.
Stellar spectroscopic surveys now have the resolution,
signal-to-noise, wavelength coverage, and sample sizes to deliver many
different chemical tags for each star.

@@

@@

The k-means algorithm has disadvantages, of course.
One is that $K$ must be chosen by hand (or heuristically at best).
Here we are only demonstrating a concept; we don't need to have the
best possible clustering.
For this reason we simply choose $K=128$, $K=256$, and $K=512$ and look at all the
results.
The algorithm also only performs local optimization.
At each $K$ we perform 32 restarts with different initializations, and
preserve the best clustering (best according to the k-means score).
Each of the 32 initializations was performed with the
\project{scikit-learn} standard k-means initialization procedure
\citep{sklearn}.

@@

@@

When the k-means results are returned, the $D\times D$ empirical
variance tensor for the members of each cluster can be constructed,
and an effective density can be computed as the number of points in
the cluster divided by the square-root of the determinant of the
tensor.
This density was used to rank abundance-space overdensities (from highest
to lowest) for visual inspection.
In \figurename~\ref{fig:densities} we show the distribution of cluster membership
and density for the $K=256$ run of k-means.
The densest clusters tend to have fewer members, presumably because
the bulk of the population is not extremely strongly clustered, or not
in ways to which k-means is sensitive.

Importantly, the k-means algorithm assigns \emph{every} star to at
least one cluster.
For this reason, there is no sense in which each ``cluster'' returned
by k-means is in any sense an overdensity in abundance space.
In what follows, we only consider high-density clusters---clusters
that are much more dense than average; these ought to represent true
over-densities in abundance space.

We chose a few interesting cases from the high-density clusters
in the $K=256$ run and show them in
\figurename~\ref{fig:M13}, \ref{fig:Pal5}, \ref{fig:Sgr},
\ref{fig:halo}, and \ref{fig:disk}.
The first three of these are dominated by stars in the known clusters
M13, Pal5, and Sagittarius.
The fourth is a halo structure with high velocity dispersion and
possibly accreted.
The fifth is a thin-disk star-formation feature.
We will discuss the astrophysical implications of these results in
\sectionname~\ref{sec:discussion}.

The abundance-space clusters shown in the \figurename s
might not look dense in abundance space, but they
are very dense in the 15-dimensional space.  Indeed,
\figurename~\ref{fig:M13} is the densest cluster found in the 15-d
space \emph{by far} (see \figurename~\ref{fig:densities}).
The challenge of this work is to find structure in the high
dimensionality space that is not visible in any 2-d projection.
Although we have by no means any model of the 15-d space, we do know
that the structures shown in the figures are very high in 15-d
density.

We have only shown results from the $K=256$ run of k-means.
This run was chosen because its densest abundance-space clusters map
well onto known stellar clusters.
At $K=128$ the densest abundance-space clusters tended to combine
multiple known stellar clusters into single, large abundance-space
groupings.
At $K=512$ k-means tended to split even mono-abundance groups into
smaller sub-groups.
This is a reminder that k-means has been chosen just for simplicity
and reliability here; it is by no means matched to the discovery of
stellar structures.
An important follow-up project is to build a model of abundance space
that captures the features expected from stellar populations.
Even if we were able to tune k-means in some sense ``perfectly'', it
would still break up many stellar clusters, since they often show
multiple stellar populations.
It is also the case that each of the abundance-space clusters we
\emph{do} show in the \figurename s includes both star-cluster members
and some background contamination, and also is missing some true
star-cluster members.

Although we show only three stellar clusters in
\figurename~\ref{fig:M13}, \ref{fig:Pal5}, and \ref{fig:Sgr}, many
other clusters are visible among our densest abundance-space k-means
clusters.
These include M15, M92, and M107, among others.
We haven't asked yet whether we detect---as abundance-space
overdensities---all of the stellar clusters we expect to find.
This \documentname\ is simply a demonstration of the chemical-tagging
concept; an investigation of whether we can construct complete
catalogs of stellar clusters is beyond the scope.

\begin{figure}[!p]
\includegraphics[width=2\figwidth]{./figs/clusters_0256.png}
\caption{The distribution of membership and density for the 256
  abundance-space clusters returned by the k-means algorithm at
  $K=256$.  The densest cluster (which is displayed in more detail in 
  \figurename~\ref{fig:M13}) is far more dense than the second
  densest.\label{fig:densities}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0253}
\caption{Same as \figurename~\ref{fig:all} except that the entire
  sample has been made gray and the members of the most dense
  $K=256$ abundance-space clusters have been rendered as unique, prominent triangles.
  This cluster, which was identified only in abundance space (six
  projections of which are the top six panels in this \figurename),
  turns out to be dominated by the halo globular cluster
  M13.\label{fig:M13}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0034}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by the disrupting globular
  cluster Palomar 5.\label{fig:Pal5}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0177}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by the Sagittarius star
  cloud.\label{fig:Sgr}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0010}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a high-velocity-dispersion
  structure in the Galaxy halo.\label{fig:halo}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0141}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a very thin young stellar
  structure in the Galaxy disk.\label{fig:disk}}
\end{figure}
